---
title: "Final_project"
format: html
editor: visual
echo: false
code-fold: true
warning: false
error: false
authors: 'Andrew Castillo, Ania Grudzien, Kayla McComb'
bibliography: ../data/references.bib
---

## Final Project

# Abstract

Imagination, often celebrated for its role in creativity and development, can also intersect with maladaptive personality traits, echoing the archetype of the 'tormented genius.' This study examines the structure of imagination, analyzing a large-scale cross-sectional dataset (N = 10,000) with a planned-missingness design. The functional sample size (pairwise-n = 8,500) was derived from the mean number of pairwise item administrations in the item pool. Our analysis found four factors emerge, which have been titled Hopelessness, Anxiety, Resentment, and Shame. These results highlight the dynamic, multi-faceted nature of imagination, underscoring its contributions to positive mental health and the development and maintenance of personality pathology.

# Introduction

Imagination is a fundamental human ability which permeates our daily experiences, influencing everything from the mundane to the grand narratives of myths and legends. Central to creativity, problem-solving, and foresight, imagination shapes our perceptions of reality and the cognitive schemas we use to interpret the world around us @gotlieb2019imagination. As a multifaceted construct, imagination involves varying levels of mental imagery, social cognition, mental simulation, emotion, and temporal exploration @abraham2016imaginative, using combinations of existing memory to create an experience distinct from external sensory stimuli. Though widely recognized as a positively valanced construct, imagination plays a complex role in mental health and personality. The tendency to frequently and intensely envision negative future scenarios is a feature of clinical depression @gotlib2010cognition and anxiety @wu2015episodic, while difficulty in distinguishing imagination from reality plays a role in thought disorders such as schizophrenia @rasmussen2022anomalies. The intersection of imagination with psychopathology highlights the challenges associated with measuring such a complex construct and evaluating its impacts on mental health.

# The Imaginative Process

Imagination is a cognitive function that encompasses various processes and is integral to human psychological functioning across a range of contexts. Abraham (2016) identifies five core categories that define imagination: perceptual/motor-related mental imagery, intentionality or recollective processing, novel combinatorial or generative processing, aesthetic phenomenology, and altered psychological states. This is extended by Crespi (2020), who asserts there are seven major components of imagination: pretend play, creativity, narratives and aesthetics, mental time travel, salience, mental imagery and sensory systems, and a neural system enabling imagination. These categories, grounded in both philosophical theory and empirical neuroscience evidence, highlight imagination's virtually limitless capacity and varied manifestations. Each category has a distinct neural basis and operates through separate cognitive systems, contributing to the multifaceted nature of imaginative cognition. Common processes, such as mental simulation @markman2012handbook, mental time travel, and perspective-taking, involve dynamic interplay between these categories. This interplay allows individuals to transcend the immediate present with novel thoughts about possible, past, and future scenarios.

Mullally & Maguire (2014) emphasize that memory forms the foundation for such imaginative projections, enabling the construction of future scenarios based on past experiences. This predictive aspect of imagination ("foresight") is crucial for adaptive functioning, often aligned with personal and social objectives through cognitive control mechanisms @Sass:01d. While the neuroscientific understanding of some aspects of imagination, such as imagery, is well-developed, others like altered states remain less explored. One key area in this ongoing exploration is the Default Mode Network @buckner2008brain. Continual research into the facets of imagination is poised to uncover new insights and relationships. However, it's crucial to recognize that dysfunction in imagination, such as excessive daydreaming @somer2002maladaptive or problematic psychological states @abraham2016imaginative can be maladaptive. This underscores the importance of imagination as both a cognitive function and an enduring source of individual differences across the lifespan.

The evolution in psychometric research towards multifactor, dimensional models reflects a significant shift in how imagination is measured. Historically, tools like the Torrance Tests of Creative Thinking @torrance1966torrance and Gough's (1979) Creative Personality Scale approached assessing imagination as a component of creativity. However, recent advancements have led to direct measures of imagination as a standalone construct. Notable advancements include the Imaginal Processes Inventory (IPI; Singer & Antrobus, 1963; Singer & Antrobus, 1966), a cross-cultural measure by Feng et al. (2017), and dimensional scales such as the Four-Factor Imagination Scale @Zabelina:2020 and the Dual-Factor Imagination Scale (DFIS; Sassenberg et al., 2023). However, the need for direct and clinically useful assessment instruments remains.

# The Current Study

This study aims to explore the structure of maladaptive imagination. Specifically, the primary objective is to develop a psychometric assessment instrument that captures the imaginative processes that inflict harm on mental health and well-being.

Examining this relationship will clarify the extent to which imagination can prove maladaptive in various contexts. For example, in affective disorders like depression, imaginative processes may contribute to persistent negative anticipatory thinking and catastrophizing, acting as a feature of its symptomatology. In thought disorders such as schizophrenia, dysfunctional imaginative processes could relate to difficulties in distinguishing between internal imaginings and external reality. In the realm of Axis II disorders, persistent imaginative patterns that result in maladaptive decision making and elevated instability in affect may underpin certain personality disorders. This confluence between imaginative processes and personality pathology highlights the potential for enhancing existing therapeutic methodologies by integrating imagination. Clinical approaches such as Cognitive Behavioral Therapy (CBT) and Art Therapy already target malformed cognitions and leverage imaginative expression as therapeutic tools @malchiodi2011handbook and could further benefit from the insights of this research by targeting specific maladaptive imaginative processes

# Hypotheses

We hypothesize there will be four distinct factors of imagination that emerge. Anxiety, defined by imagination that

# Methods

To prepare our data for analysis, we will need to simulate it. This will be achieved using rnorm() and round(). The four factors will be simulated separately in "blocks" of 25 items each and merged together. These blocks will have different, randomly selected means and standard deviations in order to allow the factor analysis to locate the desired clusters of items from each of the factors. A spattering of NAs will also be introduced using something like this: sample(1:n, 0.15nrow(data)ncol(data), replace = T) paired with a for loop. This will introduce some amount of missingness (in this case, 15%) to make the data more realistic and add a bit of intrigue and decision making around handling the missing data. After this, the data will be ready for analysis including imputation, removal of missing data, partitioning into holdouts, and more. In this stage, we will use mutate() and other tidyverse() functions per the project requirements.

```{r}
suppressMessages(suppressWarnings(library(rio)))
suppressMessages(suppressWarnings(library(here)))
suppressMessages(suppressWarnings(library(psych)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(tidyr)))
suppressMessages(suppressWarnings(library(ggplot2)))
suppressMessages(suppressWarnings(library(corrplot)))
suppressMessages(suppressWarnings(library(glmnet)))
suppressMessages(suppressWarnings(library(tidyverse)))
suppressMessages(suppressWarnings(library(MASS)))
suppressMessages(suppressWarnings(library(reshape2)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(patchwork)))
```

Data is simulated using a random sampling procedure. Four latent factors consisting of items rated 1 - 6 are produced using the sample() function and stored into a dataframe.

```{r}
set.seed(1234)

#Set parameters
n_items = 100   
n_respondents = 10000   
n_factors = 4 
noise_sd = 0.2 #Adding some noise to make this more realistic

#Generate a covariance matrix
factor_loadings = matrix(runif(n_items * n_factors), ncol = n_factors)
cov_matrix = factor_loadings %*% t(factor_loadings)

#Ensure the covariance matrix is positive definite
cov_matrix <- cov_matrix + diag(1e-6, n_items)

#Generate data
data <- mvrnorm(n = n_respondents, mu = rep(0, n_items), Sigma = cov_matrix)

#Add noise

data <- data + matrix(rnorm(n_respondents * n_items, sd = noise_sd), ncol = n_items)

data <- as.data.frame(data)
colnames(data) <- paste("Item", 1:n_items, sep = "")


#Scale the data to 1-5 range
simulated_data <- as.data.frame(lapply(data, function(x) round(4 * (x - min(x)) / (max(x) - min(x)) + 1, 0)))


gender <- sample(c("F", "M"), n_respondents, replace = TRUE, prob = c(0.66, 0.34)) #Simulating gender. 66% F/34% M as per the original sample.

ages <- pmax(pmin(round(rnorm(n_respondents, mean = 30, sd = 14)), 99), 18) #Simulating skewed distribution with a mean around 30, sd of around 14 as per the original sample.

race_categories <- c("White", "Black", "Asian", "Hispanic", "Other") #Simulating race similar to the original data
race <- sample(race_categories, n_respondents, replace = TRUE, prob = c(0.7, 0.05, 0.1, 0.1, 0.05))

countries <- c("USA", "Canada", "UK", "Germany", "Australia", "India", "China", "Brazil", "South Africa", "Japan") #Simulating location using a subset of countries (using all 200 is too taxing for this project)
geographic_location <- sample(countries, n_respondents, replace = TRUE, prob = c(0.3, 0.1, 0.1, 0.05, 0.05, 0.1, 0.1, 0.1, 0.05, 0.05))

participant_id <- sprintf("%05d", 1:n_respondents) #Giving each row a unique identifier

#Making it harder on ourselves by including missing data (the real data has a lot of missingness)
num_missing <- round(0.15 * n_respondents)

#Function to introduce missing data into a column
introduce_missing <- function(column) {
    #Randomly select indices to be set as NA
    missing_indices <- sample(1:n_respondents, num_missing, replace = FALSE)
    column[missing_indices] <- NA
    return(column)
}

#Here, we introduce 15% missingness into the data
for(i in 1:100){
  simulated_data[,i] <- introduce_missing(simulated_data[,i])
}

scale_data <- data.frame(participant_id, ages, factor(gender), factor(race), factor(geographic_location), simulated_data) #Consolidating everything into one set.

#NOTE: In the scale data, item_1 through item_25 correspond to factor 1, and so on up to factor 4.

colnames(scale_data)[1:5] <- c('participant_id', 'age', 'gender', 'race', 'country') #Cleaning up the names

write.csv(scale_data, here(file_path <- here("data", "scale_data.csv"))) #Saving it to the data folder

```

# Results

First, sample demographics and visualizations are provided below. As can be seen, the simulated sample is quite diverse in terms of number of countries represented, ages represented and, races represented.

```{r}
#comparing age demographics for white and non-white races

age1 <- scale_data %>%
filter(race == "White") %>%
summarize(mean(age))

age2 <- scale_data %>%
filter(race != "White") %>%
summarize(mean(age))

break_down<- scale_data %>%
  group_by(gender, race, country) %>%
  mutate(avg_age= mean(age))


#demographics descriptives

country <- scale_data %>%
  group_by(country) %>%
  summarise(total= n()) 

gender <- scale_data %>%
  group_by(gender) %>%
  summarise(total= n()) %>%
  mutate(frequency = total / sum(total))

race <- scale_data %>%
  group_by(race) %>%
  summarise(total= n())

gender %>%
  kable(col.names= c("Gender", "Number of Participants", "Porportion of sample"))

race %>%
  kable(col.names= c("Race", "Number of Participants"))

 country %>%
  kable(col.names= c("Country", "Number of Participants")) 
 
 age1 %>%
   kable(col.names= "Average age of White Sample", digits = 2)
 
 age2 %>%
   kable(col.names= "Average age of Non-White Sample", digits= 2)
 
 #visualizing demographics

break_down %>%
  ggplot(aes(avg_age, race))+
  geom_col(aes(fill = gender), position = "dodge") +
  facet_wrap(~country, ncol= 2) +
  theme_minimal() +
  labs(
    title = "Average Age by Race and Country",
    x = "Average Age",
    y = "Race",
    fill = "Gender"
  )
```

We found four factors of imagination emerge from our analyses. Data is visualized below. (more info)

```{r}
#Descriptives of the 100 items
descScaleItems <- psych::describe(scale_data[,6:105])

#Administrations of each
administrations <- describe(descScaleItems$n)

#Pairwise complete administrations
pwiseAdminsScaleItems <- pairwiseCount(scale_data[,6:105], diagonal = TRUE)
pwiseAdminsScaleItemsVec <- as.vector(pwiseAdminsScaleItems[lower.tri(pwiseAdminsScaleItems)])
pwiseAdminsScaleItemsVecDescribe <- psych::describe(pwiseAdminsScaleItemsVec)
n_obs <- pwiseAdminsScaleItemsVecDescribe$mean #Mean number of pairwise-complete administrations of all items

#descriptive table 
descScaleItems %>% kable(
  caption = "Descriptive Statistics of 100 Items", 
  col.names = c("Item", "Number of Observations", "Mean Response", "Standard Deviation", "Median Response", "Trimmed Mean Response", "Median Absolute Deviation", "Minimum", "Maximum", "Range", "Skew", "Kurtosis", "Standard Error" ),
  digits = 4
)




```

Cleaning the data involves the handling of missing values. 15% of the data is missing and there are no complete rows. In order to address this missingness, imptuation is used with the makeX() function from the glmnet() package.

```{r}
#Handling NAs by replacing with mean of column 

imp_sim <- data.frame(makeX(scale_data[,6:105], na.impute = TRUE, sparse = FALSE))


#creates version of data where each row is an individual score 

scale_data_long <- pivot_longer(
  data = imp_sim,
  cols = 1:100,
  names_to = "item",
  values_to = "score"
)

#creates version of data where each participant has 4 rows (one for each factor)

scale_data_wide <- scale_data_long %>% 
  mutate(item = parse_number(item)) %>% 
  mutate(factor = case_when(
    item <= 25 ~ 1,
    item > 25 & item <= 50 ~ 2,
    item > 50 & item <= 75 ~ 3,
    item > 75 ~ 4)) %>% 
  pivot_wider(
    names_from = item,
    values_from = score
  )

#creates separate dataframes for each factor 

factor_1 <- scale_data[,1:30]

factor_2 <- scale_data[,c(1:5, 31:55)]

factor_3 <- scale_data[,c(1:5, 56:80)]

factor_4 <- scale_data[,c(1:5, 81:105)]

```

The data is visualized in multiple ways, including a correlation table, heatmap, and other plots.

```{r}

#doing a correlation matrix of all the items

matrix <- cor(imp_sim, use="pairwise") 

heatmap(matrix, Rowv = NA, Colv = NA)

#making a correlation plot 
corrplot(corr= matrix, method = "square")

#Correlation matrix for factor 1
matrix1 <- cor(imp_sim[,1:25], use="pairwise")

m1 <- function() { corrplot(corr= matrix1, method = "square", tl.pos = 'n') }

#Correlation matrix for factor 2
matrix2 <- cor(imp_sim[,26:50], use="pairwise")

m2 <- function() { corrplot(corr= matrix2, method = "square", tl.pos = 'n') }

#Correlation matrix for factor 3
matrix3 <- cor(imp_sim[,51:75], use="pairwise")

m3 <- function() { corrplot(corr= matrix3, method = "square", tl.pos = 'n') }

#Correlation matrix for factor 4
matrix4 <- cor(imp_sim[,76:100], use="pairwise")

m4 <- function() { corrplot(corr= matrix4, method = "square", tl.pos = 'n') }

#Setting up the layout
par(mfrow=c(2,2), mar=c(4,4,2,2) + 0.1)


#Plotting each corrplot in the layout
m1()
m2()
m3()
m4()

#Resetting to default layout
par(mfrow=c(1,1), mar=c(5,4,2,2) + 0.1)

```

In the above four-tile correlation plot, the displayed factors clockwise from the top left are: 1) Anxiety, 2) Hopelessness, 3) Resentment, 4) Shame.

```{r}
#visualizing mean responses for each factor 

mean_scores <- scale_data_long %>% 
  group_by(item) %>% 
  summarize(mean = mean(score))

mean_scores %>% 
  mutate(item = parse_number(item)) %>% 
  mutate(factor = case_when(
    item <= 25 ~ "Anxiety",
    item > 25 & item <= 50 ~ "Hopelessness",
    item > 50 & item <= 75 ~ "Resentment",
    item > 75 ~ "Shame")) %>% 
  ggplot(aes(x = item, y = mean)) +
  geom_col(fill= "blue1") +
  labs(y = "Mean Response (1-5)",
       x = "Item",
       title = "Mean Responses to Items by Factor") +
  facet_wrap(~factor) +
  theme_minimal() 
```

```{r}
#factor analysis

fa_result <- fa(imp_sim, nfactors = n_factors, rotate = "varimax")

#extracting loadings
loadings <- as.data.frame(fa_result$loadings[, 1:n_factors])
loadings$item <- rownames(loadings)

#Sort items by their loadings on the first factor
sorted_loadings <- loadings[order(-abs(loadings[,1])), ]
sorted_items <- sorted_loadings$item

sorted_data <- imp_sim[, sorted_items]

cor_matrix <- cor(sorted_data)

reorderFA <- suppressMessages(suppressWarnings(fa(imp_sim, nfactors = 4)))
reorderItems <- rownames(fa.sort(reorderFA$loadings))
items <- reorderItems

redoneItems <- cor(imp_sim)[items,items]

var_explained <- fa_result$Vaccounted
cum_var_explained <- cumsum(var_explained[2,])

factor_plot <- corrplot(redoneItems, method = "color", 
         bg = "White", type = "full", title = "",
         col = COL2('RdBu',200), 
         tl.pos = 'n',
         #addCoef.col = 'black', 
         tl.cex = 1)
```

Lastly, we add a proportion of variance explained plot to show how well the factors account for the variance in the response data.

```{r}
#Plot
ggplot(data.frame(Factor = 1:n_factors, CumulativeVariance = cum_var_explained), aes(x = Factor, y = CumulativeVariance)) +
    geom_line() + geom_point() +
    labs(x = "Factor", y = "Cumulative Variance Explained", title = 'Proportion of Variance Explained during Factor Analysis')
```

# Discussion

In this study, we observed meaningful patterns of associations between the pool of imagination items administered, substantiating the hypothesis that imagination can prove maladaptive in various contexts and that it can be quantified in a four-factor structure. 25 items loaded on to each respective factor of Anxiety, Hopelessness, Resentment, and Shame.

These findings align with prior research @castillo2023pathology which show imaginative emotional valence is strongly associated with pathological personality traits, suggesting a clinically significant link between imagination and affective psychopathology. This is particularly the case for the emergent factors of Shame and Hopelessness, which are defined by emotionally saturated items (e.g. "I see no hope for the future" or "I feel ashamed when I picture myself in comparison to others").

Similarly, the Anxiety factor is similar to the Frequency dimension of the FFIS, revolving around persistent intrusive thoughts which impede adaptive behaviors and can result in perceptual dysregulation. The observed patterns in this study mirrored the patterns observed in Castillo & Condon (2023). These associations point to the potential of excessive engagement in imagination in contributing to thought disorders @andrews2021mapping, indicating imaginative frequency is of clinical significance. These findings reinforce the clinical relevance of imaginative frequency and anxiety induced by imagination. Of the four emergent factors, resentment demonstrated the weakest psychometric properties, which may be attributable to the factor's association with antisocial traits like the Dark Triad. Aggression and imagination are known to correlate with opposing constructs (imagination with openness, aggression with limited openness) and this may be resulting in attenuation effects. The Dark Triad literature @kay2022taking suggests a link with Machiavellianism, relating to adopting a socially predatory strategy. Our findings do support this, suggesting resentment may play into plotting revenge and attempting to set verbal "traps" for others.

# Conclusion

Overall, these results highlight the clinical potential of these imagination factors in assessing psychopathology and treating its symptoms and underlying causes. The ability of these models to predict the relationship between imagination and personality pathology points to the importance of considering imaginative processes in clinical assessments and interventions.

# References
